---
output:
  html_document:
    keep_md: yes
---
# Reproducible Research Assignment 1
**Please refer to figure folder for the figures**

The following steps are done to get started:
```{r}
# setting path to the directory containing all the files
setwd("/Users/Jinyan/RepData_PeerAssessment1/")
```

```{r, message = F, warning = F}
# loading the necessary libraries
library(dplyr)
library(ggplot2)
library(data.table)
```

```{r}
# read the .csv file, setting header to be TRUE
activity <- read.csv("activity.csv", header  = TRUE)
```

## Question 1: What is mean total number of steps taken per day?

```{r}
# grouping the data by date and summarising the resulting data table
stepsDate <- group_by(activity, date)
stepsSum <- summarise(stepsDate, sum(steps, na.rm = TRUE))
colnames(stepsSum) <- c("date", "steps")
```

```{r plot1, fig.path = "/Users/Jinyan/RepData_PeerAssessment1/figures/"}
# plotting histogram using ggplot2 system
png("plot1.png", width = 480, height = 480, type = "quartz")
g <- ggplot(stepsSum, aes(x = steps))
z <- g + geom_histogram(binwidth = max(stepsSum$steps)/30) + 
        labs(title = "Histogram of Total Steps Taken Per Day") +
        labs(x = "Total Steps Taken", y = "Frequency")
dev.off()
print(z)
```

This table shows you the summary of the total, mean and median of steps taken
per day:
```{r}
# calculating the mean and median of the steps taken per day
stepsMean <- summarise(stepsDate, mean(steps, na.rm = TRUE))
stepsMedian <- summarise(stepsDate, median(steps, na.rm = TRUE))
summaryTable <- cbind(stepsSum, stepsMean[, 2], stepsMedian[, 2])
colnames(summaryTable) <- c("Date", 
                            "Steps Taken (Total)", 
                            "Steps Taken (Mean)", 
                            "Steps Taken (Median)"
                            )
print(summaryTable)
```

## Question 2: What is the average daily activity pattern?

```{r}
# grouping the data by interval and summarising the resulting data table
stepsInt <- group_by(activity, interval)
stepsIntMean <- summarise(stepsInt, mean(steps, na.rm = TRUE))
colnames(stepsIntMean) <- c("interval", "steps")
```

```{r plot2, fig.path = "/Users/Jinyan/RepData_PeerAssessment1/figures/"}
# plotting time series plot using ggplot2 system
png("plot2.png", width = 480, height = 480, type = "quartz")
h <- ggplot(stepsIntMean, aes(interval, steps))
y <- h + geom_line() +
        labs(title = "Time Series Plot of Interval VS Average Steps Taken") +
        labs(x = "Interval", y = "Average Steps Taken") + 
        geom_vline(xintercept = 835, linetype = "longdash")
dev.off()
print(y)
```

```{r}
# determining which 5-minute interval has the highest average
stepsMax <- which(stepsIntMean$steps == max(stepsIntMean$steps))
maxSteps <- stepsIntMean$interval[stepsMax]
```
The 5-minute interval with highest number of steps taken is **`r maxSteps`** as
shown in the plot above.

## Question 3: Imputing missing values

```{r}
# total number of NAs in the data
totalNA <- sum(!is.na(activity$steps))
```
The 5-minute interval with highest number of steps taken is **`r totalNA`**.

Since on certain days, total steps taken = 0, we shall use average of steps 
taken per interval to fill in the NA values.
We will select from the mean table the respective values from the intervals.

```{r}
# converting the activity into data table for merge function later
activityDT <- data.table(activity, key = "interval")
stepsIntMeanDT <- data.table(stepsIntMean, key = "interval")
```

```{r}
# the following 3 lines convert the NA values to the average values for each
# respective intervals
mergedDT <- activityDT[stepsIntMeanDT]
mergedDT$steps <- as.double(mergedDT$steps)
mergedDT <- mergedDT[is.na(steps), steps := i.steps]
```

```{r}
# selecting the columns that you want after the conversion
# mergedDT is the new dataset with missing data filled in
mergedDT <- mergedDT[, list(steps, date, interval)]
```

```{r, warning = F}
# grouping the data by date and summarising the resulting data table
stepsDate2 <- group_by(mergedDT, date)
stepsSum2 <- summarise(stepsDate2, sum(steps))
colnames(stepsSum2) <- c("date", "steps")
```

```{r plot3, fig.path = "/Users/Jinyan/RepData_PeerAssessment1/figures/"}
# plotting histogram using ggplot2 system
png("plot3.png", width = 480, height = 480, type = "quartz")
i <- ggplot(stepsSum2, aes(x = steps))
x <- i + geom_histogram(binwidth = max(stepsSum2$steps)/30) +
        labs(title = "Histogram of Total Steps Taken Per Day (Filled NAs)") +
        labs(x = "Total Steps Taken", y = "Frequency")
dev.off()
print(x)
```

This table shows you the summary of the total, mean and median of steps taken
per day after filling in the missing values:
```{r}
stepsMean2 <- summarise(stepsDate2, mean(steps))
stepsMedian2 <- summarise(stepsDate2, median(steps))
summaryTable2 <- cbind(data.frame(stepsSum2), data.frame(stepsMean2)[, 2], 
                       data.frame(stepsMedian2)[, 2])
colnames(summaryTable2) <- c("Date", 
                            "Steps Taken (Total)", 
                            "Steps Taken (Mean)", 
                            "Steps Taken (Median)"
)
print(summaryTable2)
```
To find out the effect of filling in the missing values:
```{r}
before <- sum(summaryTable[, 2])
after <- sum(summaryTable2[, 2])
after <- as.integer(after)
```
The difference between the number of steps taken before and after filling in the
missing values is calculated by summing up the averages, giving 
**before = `r before`** and **after = `r after`**. Thus, filling in the missing 
values increases the total number of steps taken.

## Question 4: Are there difference in activity patterns between weekdays and weekends?

```{r}
# converting the dates to weekdays/weekends
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
weekdays1 <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity$wDay <- factor((weekdays(activity$date) %in% weekdays1),
                        levels = c(FALSE, TRUE), 
                        labels = c("Weekend", "Weekday"))

# grouping the data by date and weekday and summarising the resulting data table
activityWday <-group_by(activity, wDay, interval)
wdayMean <- summarise(activityWday, mean(steps, na.rm = TRUE))
colnames(wdayMean) <- c("day", "interval", "steps")
```

```{r plot4, fig.path = "/Users/Jinyan/RepData_PeerAssessment1/figures/"}
# making a panel plot of time series using ggplot2 system
png("plot4.png", width = 480, height = 480, type = "quartz")
wp <- ggplot(wdayMean, aes(interval, steps))
w <- wp + geom_line() + facet_grid(day ~ .) + 
        labs(title = "Time Series Plot of Interval VS Average Steps Taken") +
        labs(x = "Interval", y = "Average Steps Taken")
dev.off()
print(w)
```

## Thank you and congratulations on reading everything! =)